---
title: "Introduction to tooth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tooth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 5,
  dpi = 100
)
```

The **tooth** package provides tools for dental public health research:

- Caries index calculation (DMFT/dmft and DMFS/dmfs)
- Odontograph heatmap visualizations
- Tooth numbering system conversion (FDI, Universal, quadrant)

## Sample Data

The package includes a simulated dataset, `sim_exam`, with 20 patients, 7
teeth per quadrant, and 7 surfaces per tooth (5 coronal + 2 root).

```{r}
library(tooth)
data(sim_exam)
head(sim_exam)
```

## Calculating DMFT

`calc_dmft()` takes long-format surface data and returns one row per person
with Decayed (DT), Filled (FT), Missing (MT), and combined index counts.

```{r}
dmft <- calc_dmft(sim_exam)
head(dmft)
```

You can summarise the results:

```{r}
summary(dmft[, c("DT", "FT", "MT", "DMFT", "num_teeth")])
```

### Stratification

Add a grouping variable to your data and pass it via `strata`:

```{r}
# Simulate a treatment arm
sim_exam$treatment <- ifelse(
  as.numeric(gsub("P", "", sim_exam$record_id)) <= 10,
  "SDF", "ART"
)

dmft_by_arm <- calc_dmft(sim_exam, strata = "treatment")
head(dmft_by_arm)
```

### Separate Root Caries

When you have root surface data, you can tally root caries separately:
```{r}
dmft_root <- calc_dmft(sim_exam, root_lesion_col = "lesion_code")
head(dmft_root[, c("record_id", "DT", "MT", "DMFT", "RDT")])
```

`RDT` (root-decayed teeth) is counted independently from coronal `DT`.

## Calculating DMFS

`calc_dmfs()` works the same way but at the surface level:

```{r}
dmfs <- calc_dmfs(sim_exam)
head(dmfs)
```

## Odontograph Heatmaps

The odontograph displays per-tooth-surface values as a colour-coded
dental arch diagram.

### Preparing Data

Aggregate your surface data to proportions per tooth-surface:

```{r}
library(dplyr)

decay_prop <- sim_exam %>%
  filter(tooth_surface %in% c("buc", "lin", "mes", "dis", "occ")) %>%
  group_by(tooth_num, tooth_surface) %>%
  summarise(
    prop = mean(lesion_code %in% 3:6 & act == 2, na.rm = TRUE),
    .groups = "drop"
  )
head(decay_prop)
```

### Basic Odontograph

```{r fig.width=14, fig.height=5}
build_odontograph(
  data = decay_prop,
  value_col = "prop",
  teeth_per_quadrant = 7,
  title = "Active Caries Proportion by Surface",
  color_high = "#C62828",
  legend_title = "Caries\nProportion",
  surfaces = c("buc", "lin", "mes", "dis", "occ")
)
```

### With Root Surfaces

```{r fig.width=14, fig.height=5}
decay_all <- sim_exam %>%
  group_by(tooth_num, tooth_surface) %>%
  summarise(
    prop = mean(lesion_code %in% 3:6 & act == 2, na.rm = TRUE),
    .groups = "drop"
  )

build_odontograph(
  data = decay_all,
  value_col = "prop",
  teeth_per_quadrant = 7,
  title = "Active Caries — Coronal and Root Surfaces",
  color_high = "#C62828",
  legend_title = "Caries\nProportion",
  surfaces = c("buc", "lin", "mes", "dis", "occ", "rootb", "rootl")
)
```

### Stratified Odontograph

Use the `strata` argument to facet by a grouping variable:

```{r fig.width=14, fig.height=10}
decay_by_arm <- sim_exam %>%
  filter(tooth_surface %in% c("buc", "lin", "mes", "dis", "occ")) %>%
  group_by(treatment, tooth_num, tooth_surface) %>%
  summarise(
    prop = mean(lesion_code %in% 3:6 & act == 2, na.rm = TRUE),
    .groups = "drop"
  )

build_odontograph(
  data = decay_by_arm,
  value_col = "prop",
  teeth_per_quadrant = 7,
  title = "Active Caries",
  color_high = "#C62828",
  legend_title = "Caries\nProportion",
  surfaces = c("buc", "lin", "mes", "dis", "occ"),
  strata = "treatment",
  ncol = 1
)
```

### Primary Dentition

For primary teeth, use `dentition = "primary"` (5 teeth per quadrant):

```{r fig.width=12, fig.height=5}
primary_teeth <- paste0(
  rep(c("ur", "ul", "lr", "ll"), each = 5),
  rep(1:5, 4)
)
d_primary <- expand.grid(
  tooth_num = primary_teeth,
  tooth_surface = c("buc", "lin", "mes", "dis", "occ"),
  stringsAsFactors = FALSE
)
d_primary$prop <- runif(nrow(d_primary), 0, 0.5)

build_odontograph(
  data = d_primary,
  dentition = "primary",
  title = "Primary Dentition — Simulated Caries",
  color_high = "#E65100",
  surfaces = c("buc", "lin", "mes", "dis", "occ"),
  show_roots = FALSE
)
```

## Tooth Numbering Conversion

Convert between FDI (ISO 3950), Universal (ADA), and quadrant notation:

```{r}
# FDI to quadrant
tooth_convert("11", from = "fdi", to = "quadrant")

# Quadrant to Universal
tooth_convert("ur1", from = "quadrant", to = "universal")

# Multiple teeth
tooth_convert(c("11", "21", "36", "46"), from = "fdi", to = "quadrant")
```

## Tooth Configuration

Generate a layout table for any arch configuration:

```{r}
tooth_config("permanent", teeth_per_quadrant = 7)
tooth_config("primary")
```
